---
title: "Iris Dataset"
author: "Brittnany Graham"
date: "2026-02-19"
output: html_document
---

## Iris Dataset

The iris data set is a classic dataset in R, and contains the measurements of sepal length, sepal width, petal length, and petal width of three species of iris flowers

There will be a few questions to help guide you through this workshop

### General Advice

When looking at questions through the workshop, my best advice is to understand that R is a tool to make a big picture.
The question is generally the big picture.
You, as a future coder and scientist is to understand the puzzle pieces to make the big picture come to life

## General packages we will use

The tidyverse package houses multiple packages that is used frequently in RStudio, such as ggplot2, lubridate, dplyr, and purrr. This is to avoid loading multiple libraries each time 

```{r}
# install.packages("tidyverse")
library(tidyverse)

```
## Question 1

How many species of iris flowers in this data set? What are their names?

*Hint:* start with creating an object to put iris in (name it something other than iris, get creative)


```{r}

flowers <- iris

## head is to view the top 5/6 rows of a dataset 
head(flowers)

## the 'unique' function
unique(flowers$Species)

```































































































