---
title: "Iris Dataset practice"
author: "Brittnany Graham"
format: docx
editor: visual
---

## Iris Dataset

The iris data set is a classic dataset in R, and contains the measurements of sepal length, sepal width, petal length, and petal width of three species of iris flowers

There will be a few questions to help guide you through this workshop

### General Advice

When looking at questions through the workshop, my best advice is to understand that R is a tool to make a big picture.
Therefore, the question being asked is generally the big picture.
You, as a future coder and scientist is to understand the puzzle pieces to make the big picture come to life

## General packages we will use

The tidyverse package houses multiple packages that is used frequently in RStudio, such as ggplot2, lubridate, dplyr, and purrr.
This is to avoid loading multiple libraries each time

```{r}
# install.packages("tidyverse")
library(tidyverse)

# putting the iris dataset in a new object name
```

::: callout-note
## Differences between warnings and errors

Warning errors are normally harmless in RStudio.
They are simply warning you of some things that it's doing behind the scenes.
However, errors means something is critically wrong and R terminated the line of code you tried to run.
Error messages can sometimes be vague, but we can work through a few of them together.
:::

## Task One: Explore the data

Take a moment and explore the Iris data set.
Play around with different functions!
What are some returns you get?
What are some questions that first pop into your mind?
What kind of structure is the data set?

```{r}
str(flowers)
head(flowers)

```

## Question 1

How many species of iris flowers in this data set?
What are their names?

*Hint:* start with creating an object to put iris in (name it something other than iris, get creative)

```{r}

flowers <- iris

## head is to view the top 5/6 rows of a dataset 
head(flowers)

## the 'unique' function
unique(flowers$Species)

```

Answer here:

## Question 2

What is the average petal width of each species?

*Hint*: You'll want to subset each species into their own object names

**Answer code:**

```{r}

unique(flowers$Species)

# subsetting the different species
setosa <- flowers %>%
  filter(Species == "setosa")

versicolor <- flowers %>%
  filter(Species == "versicolor")

virginica <- flowers %>%
  filter (Species == "virginica")

## Next going to calculate the average petal width of each species

mean(setosa$Petal.Width)
mean(versicolor$Petal.Width)
mean(virginica$Petal.Width)
```

## Question 3

This question is in two parts, don't worry, I will help you through each step!
The beauty of R is that once you do it once, you can modify and update the code to reflect what you need for a different project!

::: {#ggplot2}
This is where ggplot2 gets introduced!
It's a package within the tidyverse package, but it's government name is "Grammar of Graphics"

It's a way to create beautiful, publish quality plots in RStudio.

Talk more here about the structure of ggplot()
:::

First, create a scatter plot that showcases the relationship between sepal length and petal length for the entire data set

-   *Don't forget* break this big picture question into small puzzle pieces

**First bit**

```{r}
global_plot <- ggplot(flowers, aes(x = Petal.Length, y = Sepal.Length, shape = Species, color = Species)) +
  geom_point() +
  theme_classic()

global_plot
```

```{r}
setosa_plot <- ggplot(setosa, aes(x = Petal.Length, y = Sepal.Length)) +
  geom_point() +
  theme_classic()

setosa_plot
```

Looks, okay, right?
You can see the points, make some general assumptions about this graph, but we can spruce it up a bit!

Let's rename the axis titles and give this plot a title for a cleaner look

```{r}
setosa_plot2 <- ggplot(setosa, aes(x = Petal.Length, y = Sepal.Length)) +
  geom_point() +
  labs(x = "Petal Length",
       y = "Sepal Length",
       title = "Setosa correlation between Petal and Sepal Length") +
  theme_classic()

setosa_plot2
```
