---
title: "Iris Dataset Practice - Answers"
author: "Brittnany Graham"
format: html
editor: visual
---

## Iris Dataset

The iris data set is a classic dataset in R, and contains the measurements of sepal length, sepal width, petal length, and petal width of three species of iris flowers

There will be a few questions to help guide you through this workshop

### General Advice

When looking at questions through the workshop, my best advice is to understand that R is a tool to make a big picture.
Therefore, the question being asked is generally the big picture.
You, as a future coder and scientist is to understand the puzzle pieces to make the big picture come to life

## General packages we will use

The tidyverse package houses multiple packages that is used frequently in RStudio, such as ggplot2, lubridate, dplyr, and purrr.
This is to avoid loading multiple libraries each time

```{r} # install.packages("tidyverse")}

library(tidyverse) 

# putting the iris dataset in a new object name
```

::: callout-tip
## Differences between warnings and errors

Warning errors are normally harmless in RStudio.
They are simply warning you of some things that it's doing behind the scenes.
However, errors means something is critically wrong and R terminated the line of code you tried to run.
Error messages can sometimes be vague, but we can work through a few of them together.
:::

### Commonly used functions in tidyverse package

Again, there are many functions and packages housed within the tidyverse package, but one of packages being dplyr, which helps the user (you) manage and provide initial summaries of datasets.
For more information, check out the following link [dplyr package](https://dplyr.tidyverse.org/reference/)

**However:** I'll provide a few functions that will help you along in the exercise below

\*`distinct()` - selects first occurrence of data

\*`rename()` - renames columns

\*`select()` - select columns

\*`filter()` - filters rows

\*`arrange()` - sort data

\*`mutate()` - adds new data/manipulations and keeps old data

\*`transmute()` - adds new data/manipulations but only keeps new data

\*`group_by()` - groups data

\*`summarize()` - summary functions

## Task One: Explore the data

Take a moment and explore the Iris data set.
Play around with different functions!
What are some returns you get?
What are some questions that first pop into your mind?
What kind of structure is the data set?

```{r}
str(flowers) 
head(flowers) 
```

## Question 1

How many species of iris flowers in this data set?
What are their names?

*Hint:* start with creating an object to put iris in (name it something other than iris, get creative)

*Challenge:* If you want to rename the column names more intuitive, how would you do that?

```{r}
flowers <- iris  

## head is to view the top 5/6 rows of a dataset  

head(flowers)  

## the 'unique' function 

unique(flowers$Species) 
```

Answer here:

1.  There are 3 species of flower: versicolor, virginica, setosa

```{r}

flowers <- flowers %>%
  rename('Sepal Length' = Sepal.Length)

flowers <- flowers %>%
  rename('Sepal Width' = Sepal.Width)

flowers <- flowers %>%
  rename('Petal Length' = Petal.Length)

flowers <- flowers %>%
  rename('Petal Width' = Petal.Width)

# Always make sure what you are executing is what you wanted the code to do. Sometimes RStudio will do exactly what you tell it to do, which is a blessing and a curse. Once you have made a change to the dataset, it becomes permanent

head(flowers)

## here you see that I was able to rename the column names of each of the variables that were measured for each species
```

## Question 2

What is the average petal width of each species?

*Hint*: You'll want to subset each species into their own object names

**Answer code:**

```{r}
unique(flowers$Species) 

# subsetting the different species 
setosa <- flowers %>%   filter(Species == "setosa")  
versicolor <- flowers %>%   filter(Species == "versicolor")  
virginica <- flowers %>%   filter (Species == "virginica")  

## Next going to calculate the average petal width of each species  

mean(setosa$`Petal Width`) 

mean(versicolor$`Petal Width`) 

mean(virginica$`Petal Width`)
```

**Answer:**

Setosa = 0.246

Versicolor = 1.326

Virginica = 2.026

## Question 3

This question is in two parts, don't worry, I will help you through each step!
The beauty of R is that once you do it once, you can modify and update the code to reflect what you need for a different project!

::: {#ggplot2}
This is where ggplot2 gets introduced!
It's a package within the tidyverse package, but it's government name is "Grammar of Graphics"

It's a way to create beautiful, publish quality plots in RStudio.

Talk more here about the structure of ggplot()
:::

First, create a scatter plot that showcases the relationship between sepal length and petal length for the entire data set

-   *Don't forget* break this big picture question into small puzzle pieces

**First bit**

```{r}

global_plot <- ggplot(flowers, aes(x = Petal.Length, y = Sepal.Length, shape = Species, color = Species)) +   
  geom_point() +   
  theme_classic()  

global_plot
```

Looks, okay, right?
You can see the points, make some general assumptions about this graph, but we can spruce it up a bit!

Let's rename the axis titles and give this plot a title for a cleaner look

```{r} global_plot2 <- ggplot(flowers, aes(x = Petal.Length, y = Sepal.Length, shape = Species, color = Species)) +   geom_point() +   labs(x = "Petal Length",        y = "Sepal Length",        title = "Petal and Sepal Length of 3 Species of Flowers") +   theme_classic()  global_plot2}
```

There are many ways you can spruce up this graph!
You can change the shapes of each species, you can even change the color, change the position of the legend, capitalize the species name within the legend, and change the scale of the x and y axis.

If you are interested in how to do that, this link [R Graphics Cookbook 2nd Edition](https://r-graphics.org/) is a wonderful tool and reference.
I use it frequently when coding and making publishable graphs.

```{r} library(help = 'datasets')}
```
